### 2025-05-05 MSc.tex
Bætti Stórhöfða við, uppfærði síðasta kaflann. Tók út * fyrir undirkafla og eitthvað. Notaði svo GPT til þess að benda mér á stafsetningarvillur og uppfærði. Held að þetta sé nokkuð langt komið.

\def\thesisyear{2025}       					% Year thesis submitted
\def\thesismonth{June}					% Month thesis submitted, e.g. "December"
\def\thesisauthor{Brynjar Geir Sigurðsson}				% Thesis authoreiningaraðferðinni
\def\thesistitle{A Neural Network Approach to Predicting Gust Factors in Complex Landscape} % Title of thesis
%\def\thesissubtitle{XXSubtitle is rarely usedXX}		% Subtitle of thesis (optional)
\def\thesisshorttitle{} 	% Optional: if title of thesis is longer than 50 characters, it would not fit on the spine (kjölur) and in this case you, need to provide a short title here for the print shop. Otherwhise: make it empty
\def\thesiscredits{60} 						% Credits awarded for the project
\def\thesissubject{Mechanical Engineering}
\def\thesiskind{M.Sc.}					% E.g. M.Sc. or Ph.D. thesis
\def\thesiskindformal{Magister Scientiarum}			% E.g. Magister Scientiarum
\def\thesisschool{School of Engineering and {Natural Sciences}}		% School
\def\thesisfaculty{Industrial Engineering, Mechanical Engineering and Computer Science}% Faculty name without "Faculty of" (gets added) 
\def\thesisaddress{Dunhagi 5}			        % Faculty or school office address street
\def\thesispostalcode{107}			                % Faculty or school office zip code
\def\thesistelephone{525 4000}					% Office telephone
%\def\thesispublisher{XX}					% Publisher (not used)
\def\thesissupervisors{Kristján Jónasson}					% Names of supervisors (split by \\ if more than one).
\def\thesisnrofsupervisors{1}					% Number of supervisors (to use "Supervisor" vs. "Supervisors")
\def\thesiscommittee{Kristján Jónasson \\ Ólafur Pétur Pálsson \\ Guðrún Nína Petersen}			% Thesis commitee must include the supervisor(s)
\def\thesisexaminer{XXNN3XX}				        % Examiner (prófdómari)
\def\thesisISBN{}           					% Thesis ISBN number (keep empty: not used anymore)
\def\thesisprinting{}						% Name of printsthop (keep empty if thesis is never printed)
%\def\thesisprinting{Háskólaprent, Fálkagata 2, 107 Reykjavík}	% Name of printsthop (keep empty if thesis is never printed)
\def\thesislicense{This thesis may not be copied in any form without author permission.} % Set license here (could also be some Creative Commons license)
%\def\thesiskeywords{Keyword1, Keyword2, Keyword3}		% Keywords (not used anywhere, hence commented out)

@@

%%%%%%%%%%% START MODIFYING HERE AGAIN %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

  \newpage % Dedication page: remove completely if you have no dedication

  \thispagestyle{empty} \mbox{}

  \vfill

  \begin{center}
    \textit{
      To all the students who made the wise decision to use \LaTeX. % Replace this by your dedication
    }
  \end{center} \vspace*{5cm}

  \vfill

%%%%%%%%%%%%%%%%%%%%%%%%% If you have no dedication: remove until here %%%%%%%%%%%%%%%%%%%%

@@

\section*{Abstract}
In many industries, whether it be windmill farming or transportation, being able to precisely predict the load caused by weather is crucial to prevent harm coming to people and machinery as well as increasing productivity. The efficiency of wind turbines in high wind speed conditions is a vital challenge in optimizing renewable energy systems. Gust predictions can be instrumental in loss prevention. If looking only at wind, the most destructive moments are gusts, by definition. To predict these values there are two ways, traditional numerical weather prediction systems and machine learning or other pattern recognition systems. Computer aided numerical weather prediction systems originated in the 1950's, while the use of machine learning in numerical weather predictions is much more recent, having truly gained traction in the last decade. The goal of this thesis is try to combine these two methods to try and predict wind gust factor, where the gust factor is defined as the ratio between wind gust and average wind speed. This is done by having the input data to the machine learning model be reanalysis variables that were generated by traditional numerical weather prediction systems. In addition to these variables and derived variables, a digital elevation model (DEM) data was used. A deep neural network was constructed using some set of variables and the results are compared to a baseline as well as each other so as to show the impact each feature has. Neural networks can help make wind gust predictions in complex ladscape.
\vfill \vspace*{1cm}

\section*{Útdráttur}
Hvort sem um er að ræða vindmyllubúgarða eða flutningar er spágetan um álag vegna vinds mikilvæg til að koma í veg fyrir tjón á iðnviðum auk þess er hægt að auka framleiðni. Hviðuspár geta verið stór þáttur í tjónaminnkun. Ef einungis er horft á vind, þá eru hviður hæsti álagspunktur út frá skilgreiningu. Sögulega hefur verið notast við hefðbundin spálíkön sem reiða sig á töluleg eðlisfræðileg líkön til að spá fyrir um niðurstöður. Á síðustu árum hefur gervigreind þróast mikið og líkön verið þjálfuð til þess að giska á veður. Svo miklar framfarir hafa orðið að sum líkön geta keppt við hefðbundin veðurlíkön. Markmið þessa verkefnis er reyna að nýta bæði hefðbundin veðurlíkön og gervigreind til þess að giska á hviðustuðla, þar sem hviðustuðullinn er skilgreindur sem hlutfall hæstu hviðu og meðalvinds. Þetta er gert með því að nota endurgreiningargögn úr hefðbundnu veðurlíkani sem grunngögn í gervigreindarlíkan. Einnig eru notað hæðarlíkan til að lýsa umhverfi. Djúpt tauganet var búið til með því að velja breytur úr endurgreiningargögnunum ásamt hæðarpunktum og afleiddum breytum. Þetta líkan er svo þjálfað og niðurstöður bornar saman við grunnlíkan sem og önnur líkön með öðrum breytum til að skoða áhrif breyta. Djúp tauganet geta búið til hjálplegar spár fyrir vindhviður í flóknu landslagi.
\vfill

@@

\chapter*{Abbreviations}
\addcontentsline{toc}{chapter}{Abbreviations}
\markboth{Abbreviations}{Abbreviations}
\begin{acronym}[GeoTIFF]
  \acro{API}{Application Programming Interface}
  \acro{ASL}{Above Sea Level}
  \acro{AWS}{Automatic Weather Stations}
  \acro{AWSL}{Average Wind Speed Limit}
  \acro{CARRA}{Copernicus Artic Regional ReAnalysis dataset}
  \acro{CNN}{Convolutional Neural Networks}
  \acro{DEM}{Digital Elevaiton Model}
  \acro{ELI5}{Explain Like I am 5}
  \acro{ECMWF}{European Centre for Medium-Range Weather Forecast}
  \acro{GCM}{General Circulation Model}
  \acro{GeoTIFF}{Georeferenced TIFF}
  \acro{GPU}{Graphical Processing Unit}
  \acro{HRES}{High Resolution forecas}
  \acro{IMO}{Icelandic Meteoroligcal Office}
  \acro{IRCA}{Icelandic Road and Coastal Administration}
  \acro{JNWPU}{Joint Numerical Weather Prediction Unit}
  \acro{LWM}{Large AI Weather forecast Model}
  \acro{MAE}{Mean Asbolute Error}
  \acro{MAPE}{Mean Absolute Percentage Error}
  \acro{NN}{Neural Network}
  \acro{NWP}{Numerical Weather Prediction}
  \acro{SENS}{School of Engineering and Natural Sciences}
  \acro{TIFF}{Tag Image File Format}
  \acro{UoI}{University of Iceland}
\end{acronym}

>>>>

\def\thesisyear{2025}       					% Year thesis submitted
\def\thesismonth{June}					% Month thesis submitted, e.g. "December"
\def\thesisauthor{Brynjar Geir Sigurðsson}				% Thesis authoreiningaraðferðinni
\def\thesistitle{A Neural Network Approach to Predicting Gust Factors in Complex Landscape} % Title of thesis
%\def\thesissubtitle{XXSubtitle is rarely usedXX}		% Subtitle of thesis (optional)
\def\thesisshorttitle{} 	% Optional: if title of thesis is longer than 50 characters, it would not fit on the spine (kjölur) and in this case you, need to provide a short title here for the print shop. Otherwhise: make it empty
\def\thesiscredits{60} 						% Credits awarded for the project
\def\thesissubject{Mechanical Engineering}
\def\thesiskind{M.Sc.}					% E.g. M.Sc. or Ph.D. thesis
\def\thesiskindformal{Magister Scientiarum}			% E.g. Magister Scientiarum
\def\thesisschool{School of Engineering and {Natural Sciences}}		% School
\def\thesisfaculty{Industrial Engineering, Mechanical Engineering and Computer Science}% Faculty name without "Faculty of" (gets added) 
\def\thesisaddress{Dunhagi 5}			        % Faculty or school office address street
\def\thesispostalcode{107}			                % Faculty or school office zip code
\def\thesistelephone{525 4000}					% Office telephone
%\def\thesispublisher{XX}					% Publisher (not used)
\def\thesissupervisors{Kristján Jónasson}					% Names of supervisors (split by \\ if more than one).
\def\thesisnrofsupervisors{1}					% Number of supervisors (to use "Supervisor" vs. "Supervisors")
\def\thesiscommittee{Kristján Jónasson \\ Ólafur Pétur Pálsson \\ Guðrún Nína Petersen}			% Thesis commitee must include the supervisor(s)
\def\thesisexaminer{Halldór Björnsson}				        % Examiner (prófdómari)
\def\thesisISBN{}           					% Thesis ISBN number (keep empty: not used anymore)
\def\thesisprinting{}						% Name of printsthop (keep empty if thesis is never printed)
%\def\thesisprinting{Háskólaprent, Fálkagata 2, 107 Reykjavík}	% Name of printsthop (keep empty if thesis is never printed)
\def\thesislicense{This thesis may not be copied in any form without author permission.} % Set license here (could also be some Creative Commons license)
%\def\thesiskeywords{Keyword1, Keyword2, Keyword3}		% Keywords (not used anywhere, hence commented out)

@@

@@

\section*{Abstract}
In many industries, whether it be windmill farming or transportation, being able to precisely predict the load caused by weather is crucial to prevent harm coming to people and machinery as well as increasing productivity. The efficiency of wind turbines in high wind speed conditions is a vital challenge in optimizing renewable energy systems. Gust predictions can be instrumental in loss prevention. If looking only at wind, the most destructive moments are gusts, by definition. To predict these values there are two ways, traditional numerical weather prediction systems and machine learning (or other pattern recognition) methods. Computer-aided numerical weather prediction systems originated in the 1950s, while the use of machine learning in numerical weather predictions is much more recent, having gained traction in the last decade. The goal of this thesis is to combine these two methods to predict the wind gust factor, defined as the ratio between wind gust and average wind speed. This is done by having the input data to the machine learning model be reanalysis variables that were generated by traditional numerical weather prediction systems. In addition to these variables (and derived variables), a digital elevation model (DEM) was used. A deep neural network was constructed using some set of variables and the results are compared to a baseline as well as each other so as to show the impact each feature has. Neural networks can help make wind gust predictions in complex ladscape.
\vfill \vspace*{1cm}

\section*{Útdráttur}
Hvort sem um er að ræða vindmyllubúgarða eða stöðuleiki flutningabíla er spágetan um álag vegna vinds mikilvæg til að koma í veg fyrir tjón á iðnviðum, auk þess er hægt að auka framleiðni. Hviðuspár geta verið stór þáttur í tjónaminnkun. Ef einungis er horft á vind, þá eru hviður hæsti álagspunktur samkvæmt skilgreiningu. Sögulega hefur verið notast við hefðbundin spálíkön sem reiða sig á töluleg eðlisfræðileg líkön til að spá fyrir um niðurstöður. Á síðustu árum hefur gervigreind þróast mikið og líkön verið þjálfuð til þess að spá fyrir um veður. Svo miklar framfarir hafa orðið að sum líkön geta keppt við hefðbundin veðurlíkön. Markmið þessa verkefnis er reyna að nýta bæði hefðbundin veðurlíkön og gervigreind til þess að giska á hviðustuðla, þar sem hviðustuðullinn er skilgreindur sem hlutfall hæstu hviðu og meðalvinds. Þetta er gert með því að nota endurgreiningargögn úr hefðbundnu veðurlíkani sem grunngögn í gervigreindarlíkan. Einnig eru notað hæðarlíkan til að lýsa umhverfi. Djúpt tauganet var búið til með því að velja breytur úr endurgreiningargögnunum ásamt hæðarpunktum og afleiddum breytum. Þetta líkan er svo þjálfað og niðurstöður bornar saman við grunnlíkan sem og önnur líkön með öðrum breytum til að skoða áhrif breyta. Djúp tauganet geta búið til hjálplegar spár fyrir vindhviður í flóknu landslagi.
\vfill

@@

\chapter*{Abbreviations}
\addcontentsline{toc}{chapter}{Abbreviations}
\markboth{Abbreviations}{Abbreviations}
\begin{acronym}[GeoTIFF]
  \acro{API}{Application Programming Interface}
  \acro{ASL}{Above Sea Level}
  \acro{AWS}{Automatic Weather Stations}
  \acro{CARRA}{Copernicus Artic Regional ReAnalysis dataset}
  \acro{CNN}{Convolutional Neural Networks}
  \acro{DEM}{Digital Elevaiton Model}
  \acro{ELI5}{Explain Like I am 5}
  \acro{ECMWF}{European Centre for Medium-Range Weather Forecast}
  \acro{GCM}{General Circulation Model}
  \acro{GeoTIFF}{Georeferenced TIFF}
  \acro{GPU}{Graphical Processing Unit}
  \acro{HRES}{High Resolution forecas}
  \acro{IMO}{Icelandic Meteoroligcal Office}
  \acro{IRCA}{Icelandic Road and Coastal Administration}
  \acro{JNWPU}{Joint Numerical Weather Prediction Unit}
  \acro{LWM}{Large AI Weather forecast Model}
  \acro{MAE}{Mean Asbolute Error}
  \acro{MAPE}{Mean Absolute Percentage Error}
  \acro{NN}{Neural Network}
  \acro{NWP}{Numerical Weather Prediction}
  \acro{SENS}{School of Engineering and Natural Sciences}
  \acro{TIFF}{Tag Image File Format}
  \acro{UoI}{University of Iceland}
\end{acronym}
